---
title: "Lab_16_Jay_English"
author: "Jay English"
date: "2024-04-30"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: spacelab
    highlight: breezedark
---

```{r}
library(tidyverse)
library(knitr)
library(ggtree)
library(TDbook) 
library(ggimage)
library(rphylopic)
library(treeio)
library(tidytree)
library(ape)
library(TreeTools)
library(phytools)
library(ggnewscale)
library(ggtreeExtra)
library(ggstar)
library(ggeasy)
```

# Reading in Files

Load the MAG table with the new anotations.

```{r}
NEON_MAGs <- read_csv("data/GOLD_Study_ID_Gs0161344_NEON_2024_4_21.csv") %>% 
  # remove columns that are not needed for data analysis
  select(-c(`GOLD Study ID`, `Bin Methods`, `Created By`, `Date Added`, `Bin Lineage`)) %>% 
  # create a new column with the Assembly Type
  mutate("Assembly Type" = case_when(`Genome Name` == "NEON combined assembly" ~ `Genome Name`,
                            TRUE ~ "Individual")) %>% 
  mutate_at("Assembly Type", str_replace, "NEON combined assembly", "Combined") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "d__", "") %>%  
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "p__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "c__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "o__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "f__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "g__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "s__", "") %>%
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";", remove = FALSE) %>% 
  mutate_at("Domain", na_if,"") %>% 
  mutate_at("Phylum", na_if,"") %>% 
  mutate_at("Class", na_if,"") %>% 
  mutate_at("Order", na_if,"") %>% 
  mutate_at("Family", na_if,"") %>% 
  mutate_at("Genus", na_if,"") %>% 
  mutate_at("Species", na_if,"") %>% 
  
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "S-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

# NEON Metagenome Information

Remove the re-annotation and WREF plot samples from the data set previously exported from IMG

```{r}
NEON_metagenomes <- read_tsv("data/exported_img_data.tsv") %>% 
  select(-c(`Domain`, `Sequencing Status`, `Sequencing Center`)) %>% 
  rename(`Genome Name` = `Genome Name / Sample Name`) %>% 
  filter(str_detect(`Genome Name`, 're-annotation', negate = T)) %>% 
  filter(str_detect(`Genome Name`, 'WREF plot', negate = T)) 
```

Now reformat Genome Name as we did for the above MAG table.

```{r}
NEON_metagenomes <- NEON_metagenomes %>% 
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-")
```

# NEON Chemistry Data

```{r}
NEON_chemistry <- read_tsv("data/neon_plot_soilChem1_metadata.tsv") %>% 
  # remove -COMP from genomicsSampleID
  mutate_at("genomicsSampleID", str_replace, "-COMP", "") 
```

# Join the NEON MAG, metagenome, and chemistry dataframes

```{r}
NEON_MAGs_metagenomes_chemistry <- NEON_MAGs %>% 
  left_join(NEON_metagenomes, by = "Sample Name") %>% 
  left_join(NEON_chemistry, by = c("Sample Name" = "genomicsSampleID")) %>% 
  rename("label" = "Bin ID")
```

# Reading the tree files

```{r}
tree_arc <- read.tree("data/gtdbtk.ar53.decorated.tree")
tree_bac <- read.tree("data/gtdbtk.bac120.decorated.tree")
```

# Getting your subtree

You need to find the internal node on the tree corresponding to your phylum ("Actinobacteriota" was renamed to "Actinomycetota")

```{r}
# Make a vector with the internal node labels
node_vector_bac = c(tree_bac$tip.label,tree_bac$node.label)

# Search for your Phylum or Class to get the node
grep("Actinomycetota", node_vector_bac, value = TRUE)
```

```{r}
match(grep("Actinomycetota", node_vector_bac, value = TRUE), node_vector_bac)
```

```{r}

# First need to preorder tree before extracting. 
tree_bac_preorder <- Preorder(tree_bac)
tree_Actinomycetota <- Subtree(tree_bac_preorder, 1789)
```

# Filter NEON table to taxonomic group or site

```{r}
NEON_MAGs_Actinomycetota <- NEON_MAGs_metagenomes_chemistry %>% 
  filter(Phylum == "Actinomycetota")
```


```{r, fig.height= 50, fig.width=50}
ggtree(tree_Actinomycetota) +
  geom_tiplab(size = 3) +
  xlim(0, 25)
```

```{r, fig.height= 50, fig.width=50}
ggtree(tree_Actinomycetota, layout="circular") + 
  geom_tiplab(aes(angle=angle))+
    theme_tree() +
    xlim(0,25)
```

Circular with taxonomic group highlighted

```{r}
ggtree(tree_bac, layout="circular") +
    geom_hilight(node=1789, fill="steelblue", alpha=.6) 
```

Circular with no branch lengths taxonomic group labeled

```{r}
ggtree(tree_bac, layout="circular", branch.length="none") +
    geom_hilight(node=1789, fill="steelblue", alpha=.6) +
    geom_cladelab(node=1789, label="          Actinomycetota", align=TRUE,  
                  offset = 1, textcolor='steelblue', barcolor='steelblue')
```

Circular no branch lengths taxonomic group collapsed

```{r}
ggtree(tree_bac, layout="circular", branch.length="none") %>% 
  collapse(node=1789) + 
  geom_point2(aes(subset=(node==70)), shape=23, size=5, fill='steelblue') +
  geom_cladelab(node=55, label="Actinomycetota", align=TRUE,  
                  offset = 2, textcolor='steelblue')
```

Circular no branch lengths taxonomic group collapsed as trianle and labeled

```{r}
p <- ggtree(tree_bac, layout="circular", branch.length="none")
scaleClade(p, 1789, .7) %>% collapse(1789, 'min', fill="steelblue")  +
  geom_cladelab(node=1789, label="Actinomycetota", align=TRUE,  
                  offset = 0.2, textcolor='steelblue')
```

# Adding data to your tree

Note the special symbol %<+% is used for joining the tree object with your data.

```{r, fig.height= 50, fig.width=50}
ggtree(tree_Actinomycetota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=2, hjust=-.1) +
  xlim(0,25) +
  geom_point(mapping=aes(color=Class)) +
  easy_plot_legend_size(40) +
  easy_plot_legend_title_size(40) 
```

```{r, fig.height=50, fig.width=50}
ggtree(tree_Actinomycetota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=4, hjust=-.1) +
  xlim(0,25) +
  geom_point(mapping=aes(color=Class, shape = `Assembly Type`)) +
  easy_plot_legend_size(40) +
  easy_plot_legend_title_size(40) 
```



# Add a gradient to your tree

```{r, fig.height=50, fig.width=50}
ggtree(tree_Actinomycetota)  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tippoint(aes(colour=`Bin Completeness`)) + 
  scale_colour_gradient(low='blue', high='red') +
  geom_tiplab(size=3, hjust=-.1) +
  xlim(0,25) +
  easy_plot_legend_size(10) +
  easy_plot_legend_title_size(40)
```

# Circular tree with data

```{r, fig.height=25, fig.width=25}
ggtree(tree_Actinomycetota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_point2(mapping=aes(color=`Ecosystem Subtype`, size=`Total Number of Bases`))
```

# Multiple graphs in the same panel

```{r, fig.height=25, fig.width=25}
# For unknown reasons the following does not like blank spaces in the names
NEON_MAGs_metagenomes_chemistry_noblank <- NEON_MAGs_metagenomes_chemistry %>% 
  rename("AssemblyType" = "Assembly Type") %>% 
  rename("BinCompleteness" = "Bin Completeness") %>% 
  rename("BinContamination" = "Bin Contamination") %>% 
  rename("TotalNumberofBases" = "Total Number of Bases") %>% 
  rename("EcosystemSubtype" = "Ecosystem Subtype")

ggtree(tree_Actinomycetota)  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tippoint(aes(colour=`Ecosystem Subtype`)) + 

# For unknown reasons the following does not like blank spaces in the names
  geom_facet(panel = "Bin Completeness", data = NEON_MAGs_metagenomes_chemistry_noblank, geom = geom_point, 
      mapping=aes(x = BinCompleteness)) +
  geom_facet(panel = "Bin Contamination", data = NEON_MAGs_metagenomes_chemistry_noblank, geom = geom_col, 
                aes(x = BinContamination), orientation = 'y', width = .6) +
  theme_tree2(legend.position=c(.1, .7))
```

# Adding outer rings to a circular tree

```{r, fig.height=100, fig.width=100}
ggtree(tree_Actinomycetota, layout="circular", branch.length="none") %<+% 
  NEON_MAGs_metagenomes_chemistry + 
  geom_point2(mapping=aes(color=`Ecosystem Subtype`, size=`Total Number of Bases`)) + 
  new_scale_fill() + 
  geom_fruit(
      data=NEON_MAGs_metagenomes_chemistry_noblank,
      geom=geom_tile,
      mapping=aes(y=label, x=1, fill = AssemblyType),
      offset=0.08, pwidth=0.25) +
  easy_plot_legend_size(40) +
  easy_plot_legend_title_size(40)
```

Adding multiple rings to a circular tree

```{r, fig.height=25, fig.width=25}
ggtree(tree_Actinomycetota, layout="circular", branch.length="none") %<+% 
  NEON_MAGs_metagenomes_chemistry + 
  geom_point2(mapping=aes(color=`Ecosystem Subtype`, size=`Total Number of Bases`)) + 
  new_scale_fill() + 
  geom_fruit(
      data=NEON_MAGs_metagenomes_chemistry_noblank,
      geom=geom_tile,
      mapping=aes(y=label, x=1, fill= AssemblyType),
      offset=0.08,   # The distance between external layers, default is 0.03 times of x range of tree.
      pwidth=0.25 # width of the external layer, default is 0.2 times of x range of tree.
      ) + 
  new_scale_fill() +
  geom_fruit(
          data=NEON_MAGs_metagenomes_chemistry_noblank,
          geom=geom_col,
          mapping=aes(y=label, x=TotalNumberofBases),  
          pwidth=0.4,
          axis.params=list(
                          axis="x", # add axis text of the layer.
                          text.angle=-45, # the text size of axis.
                          hjust=0  # adjust the horizontal position of text of axis.
                      ),
          grid.params=list() # add the grid line of the external bar plot.
      ) + 
      theme(#legend.position=c(0.96, 0.5), # the position of legend.
          legend.background=element_rect(fill=NA), # the background of legend.
          legend.title=element_text(size=7), # the title size of legend.
          legend.text=element_text(size=6), # the text size of legend.
          legend.spacing.y = unit(0.02, "cm")  # the distance of legends (y orientation).
      ) +
  easy_plot_legend_size(20) +
  easy_plot_legend_title_size(20)
```

